---
title: "Module 3: R"
subtitle: "Programming"
author:
- "Instructor: Anjali Silva, PhD"
- "TA: Tia Harrison, MSc"
institute: "Data Sciences Institute, University of Toronto"
date: "7 July 2022"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  colors = c(
    red = "#f34213",
    purple = "#3e2f5b",
    orange = "#ff8811",
    green = "#136f63",
    white = "#FFFFFF"
  )
)

```

```{r, include=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)
library(kableExtra)
library(rgdal)
```

# Course Documents
* Visit: https://github.com/anjalisilva/IntroductionToR 

* All course material will be available via IntroductionToR GitHub repository (https://github.com/anjalisilva/IntroductionToR). Folder structure is as follows:
   * Lessons - All files: This folder contains all files.
   * **Lessons - Data only**: This folder contains data only.
   * **Lessons - Lesson Plans only**: This folder contains lesson plans only.
   * **Lessons - PDF only**: This folder contains slide PDFs only.
   * README - README file
   * .gitignore - Files to ignore specified by instructor

## Course Contacts
* Instructor: Anjali Silva
  Email: a.silva@utoronto.ca (Must use the subject line DSI-IntroR.   E.g., DSI-IntroR: Inquiry about Lecture I.)

* TA: Tia Harrison
Email: tia.harrison@mail.utoronto.ca

---

# Overview

- Functions (Wickham and Grolemund, 2017, Chapter 19)
- Loops (Wickham and Grolemund, 2017, Chapter 21)
- if/else logic (Alexander (eds), 2021, Chapter 47)
- purr
- Simulation (Alexander (eds), 2021, Chapter 47)

---

class: inverse, center, middle

# Functions

---

## Introduction

You can write your own functions in R, and you should consider doing so when you have copy-pasted a chunk of code twice.

---

## Structure

You provide a name, inputs (also known as arguments), and the body of the function that performs the operation.

```{r, eval=F}
function_name <- function(inputs) {
  <calculations using inputs>
  return(outputs)
}
```

When naming, try not to use names that already have meaning in R.

---

class: inverse, center, middle

# Loops

---

## Basic form

Loops are another tool for reducing the need to duplicate code, this time by repeatedly performing a task.

1.  For loops iterate over a set amount:

```{r,eval=F}
for (sequence to iterate over) {
  <code to execute>
}
```

2. While loops iterate based on a stopping condition:

```{r,eval=F}
while (iterator condition) {
  <code to execute>
}
```

---

## For loop example

```{r}
for (i in 1:10){
  print(i*5)
}
```

---

## For loops to modify an existing object

To create new column that adds the Sepal.Length in each row with the Sepal.Length from the previous row:

```{r}
for (i in 2:nrow(iris)) {
  iris$previous_combo[i] <- iris$Sepal.Length[i] + iris$Sepal.Length[i-1]
}
iris
```

---

## Different ways to loop:

You can loop over elements:

```{r}
for (i in c("a", "b", "c")){
  print(i)
}
```

You can loop over numeric indices:

```{r}
for (i in 1:3) {
  print(now( ) + i)
}
```

---

## Using a vector to collect outputs

```{r}
outputs <- c()

for (i in 1:5) {
  outputs <- c(outputs, i) * i
}

outputs
```

---

## While loop example

Note that we initiate the iterator i outside the loop and increment it in the loop. If the iterator never increases in the loop, then the loop will never end.

```{r}
i = 1
while(i <= 10){
  print(i*5)
  i = i + 1
}
```

---

class: inverse, center, middle

# If/else Logic

---

## Basic structure

```{r,eval=F}
if(condition1) {
  <code to execute if condition1 is TRUE>
} elif (condition2) {
  <code to execute if condition1 is FALSE and conditions2 is TRUE>
} else {
  <code to execute if condition1 and condition2 are both FALSE>
}
```

---

## Conditions

Conditions must either evaluate to TRUE or FALSE.

You can combine multiple conditions using the 'or' operator:

- (condition1) || (condition2)

You can combine multiple conditions using the 'and' operator:

- (condition1) && (condition2)

To find out if any of a list of conditions is TRUE, use any().

To find out if all of a list of conditions is TRUE, use all().

---

## if else function

The function if_else writes out a conditional statement in one line.

```{r, eval=F}
if_else(condition, output if condition is TRUE, output if condition is FALSE)
```

---

## Case when

When you have a list of possible conditions, you can use case_when instead.

```{r,eval=F}
case_when(condition1 ~ output1,
          condition2 ~ output2,
          condition3 ~ output3,
          ...
          )
```

---

## Example

```{r}
grades <- tibble(grade = c(94, 87, 73))

grades %>%
  mutate(letter = case_when(grade >= 80 ~ "A",
                            grade >= 70 ~ "B",
                            TRUE ~ "F"
                            ))
```

Note that each condition is checked in order: if condition1 is TRUE, output1 will be chosen and condition2 will not be checked.

---

# purr

Iteration is made more straightforward with the purr library.

---

## Mapping functions

Each type of output has a different function:
- map() for lists
- map_lgl() for logical vectors
- map_int() for integer vectors
- map_dbl() for double vectors
- map_chr() for character vectors

---

## Looping over columns in a dataset

```{r}
iris %>%
  map_dbl(mean)
```

```{r}
iris %>%
  map_chr(typeof)
```

---
## Looping over columns in a dataset

```{r}
iris %>%
  map(summary)
```

---

## Mapping over multiple arguments

```{r}
x <- list(1, 1, 1)
y <- list(10, 20, 30)

map2(x, y, ~ .x + .y)
```

---

class: inverse, center, middle

# Simulation

---
# Simulation
We can generate random data in R.

```{r}
runif(5)
runif(5)
```

The outcomes will be different every time.

---
# Simulation
If you want the results to be consistent, you must set a seed. The seed can be any number.

```{r}
set.seed(1818)
runif(5)
```

```{r}
set.seed(1838)
runif(5)
```

```{r}
set.seed(1818)
runif(5)
```

---

## The uniform distribution

```{r, eval=F}
runif(number, min, max)
```

```{r}
set.seed(1818)
runif(10, 1, 20)
```

---

## The normal distribution

```{r, eval=F}
rnorm(number, mean, sd)
```

```{r}
set.seed(1818)
rnorm(10, 5, 1)
```

---

## Sampling

```{r,eval=F}
sample(thing to sample from, size = number, replace, prob = vector of probability weights)
```

```{r}
set.seed(1818)
sample(c("a", "b", "c"),
       size = 10,
       replace = TRUE,
       prob = c(0.1, 0.2, 0.7))
```

The probability weights are optional. If you do not specify, all the results will be equally probable.

If you specify replace = FALSE, there must be as many or more in the thing that you sample from as the desired sample size.

---

## Simulating datasets

We can put our randomization skills to use and create toy datasets.

```{r}
set.seed(1818)
simulated_data <- tibble(X = runif(10, 0, 20),
       Y = 3*X + rnorm(10, 0, 1)) 

simulated_data
```
---
## Simulating datasets

```{r}
simulated_data %>%
  ggplot(aes(x = X, y = Y)) +
  geom_point()
```

---

class: inverse, center, middle

# Exercises

---
# Exercises

1-Write a greeting function that says “good morning”, “good afternoon”, or “good evening”, depending on the time of day.

2-Simulate a dataset using a normal distribution with mean 100 and standard deviation 15 as variable X, and a quadratic transformation of X as variable Y. Graph your data.

---

class: inverse, center, middle

# Any questions?


