---
title: "Module 3: R"
subtitle: "Inequity"
author:
- "Instructor: Anjali Silva, PhD"
institute: "Data Sciences Institute, University of Toronto"
date: "2022"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  colors = c(
    red = "#f34213",
    purple = "#3e2f5b",
    orange = "#ff8811",
    green = "#136f63",
    white = "#FFFFFF"
  )
)

```

```{r, include=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)
library(kableExtra)
library(rgdal)
```

# Course Documents
* Visit: https://github.com/anjalisilva/IntroductionToR 

* All course material will be available via IntroductionToR GitHub repository (https://github.com/anjalisilva/IntroductionToR). Folder structure is as follows:
   * Lessons - All files: This folder contains all files.
   * **Lessons - Data only**: This folder contains data only.
   * **Lessons - Lesson Plans only**: This folder contains lesson plans only.
   * **Lessons - PDF only**: This folder contains slide PDFs only.
   * README - README file
   * .gitignore - Files to ignore specified by instructor

## Course Contacts
* Instructor: Anjali Silva
  Email: a.silva@utoronto.ca (Must use the subject line DSI-IntroR.   E.g., DSI-IntroR: Inquiry about Lecture I.)

* TA: see GitHub

---

# Overview

- Equity, Diversity & Inclusion (EDI) 
- Stereotypes
- Bias 

---

class: inverse, center, middle

# What is EDI?

---

## Diversity

Diversity refers to differences, traits that impact on our perspectives. These include our gender, ethnicity, race, abilities, sexual orientation, age, education, values, beliefs, hobbies, family status, religion and profession.

## Inclusion

Inclusion refers to being respected, valued, welcomed and appreciated for differences.

## Equity

Even if everyone is treated equally, discrimination can occur. Equity is created when all people have access to what they need to succeed.

---

## Stereotypes

Stereotypes are overgeneralized and often inaccurate beliefs about groups of people. We usually learn them from our family, friends, school, and media.

## Bias

A bias is a personal preference, a like or dislike, that interferes with our ability to be impartial or objective.

Unconscious bias happens automatically, when our brains make quick judgments and assessments of people and situations based on previously stored information, including stereotypes.

- Appearance bias: when someone is treated differently based on how they look
- Similarity bias: when we favor people that are like us
- Affinity bias: when we connect with people who have similar experiences
- Confirmation bias: occurs when you interpret information in a way that affirms your prior beliefs or biases and ignores evidence that contradicts them.

---

## EDI training

EDI training is mandatory at many workplaces, often consisting of the definitions given previously as well as a link to the particular organization's culture, policy, or legal context.

Employees may be quizzed on the content or asked to reflect on their own positionality and intent going forward.

---


class: inverse, center, middle

# EDI in Data Science

---
## Video: Creating diverse and equitable initiatives in data science | Tiffany Oliver | TEDxMorehouseCollege

- Watch: https://www.youtube.com/watch?v=D_0cGEzgrFU

<video width="320" height="240">
<source src="https://www.youtube.com/watch?v=D_0cGEzgrFU" type="video/mp4">
</video>

---
## Video: How can Academic Data Science Alliance (ADSA) Advance Diversity, Equity, and Inclusion in Data Science?

- Watch: https://www.youtube.com/watch?v=J2oJf5f2Weg


<video width="320" height="240">
<source src="https://www.youtube.com/watch?v=J2oJf5f2Weg" type="video/mp4">
</video>

---

## Discussion questions

1. what has your previous experience of EDI training been?
2. As beginning data scientists, what impact can you have on EDI in data science? What is within your control?
3. What EDI concerns do you think are more prominent in data science than other professions?
4. What would you like to see in EDI programming for data science?

