---
title: "Module 3: R"
subtitle: "Manipulation"
author:
- "Instructor: Anjali Silva, PhD"
- "TA: Tia Harrison, MSc"
institute: "Data Sciences Institute, University of Toronto"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  colors = c(
    red = "#f34213",
    purple = "#3e2f5b",
    orange = "#ff8811",
    green = "#136f63",
    white = "#FFFFFF"
  )
)

```

```{r, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  message = FALSE, # hide messages from code in output
  warning = FALSE  # hide warnings from code in output
  )
library(tidyverse)
library(lubridate)
library(janitor)
library(kableExtra)
library(rgdal)
```

# Course Documents
* Visit: https://github.com/anjalisilva/IntroductionToR 

* All course material will be available via IntroductionToR GitHub repository (https://github.com/anjalisilva/IntroductionToR). Folder structure is as follows:
   * Lessons - All files*: These folders contain all files.
   * **Lessons - Data only**: This folder contains data only.
   * **Lessons - Lesson Plans only**: This folder contains lesson plans only.
   * **Lessons - PDF only**: This folder contains slide PDFs only.
   * README - README file
   * .gitignore - Files to ignore specified by instructor

## Course Contacts
* Instructor: Anjali Silva
  Email: a.silva@utoronto.ca (Must use the subject line DSI-IntroR.   E.g., DSI-IntroR: Inquiry about Lecture I.)

* TA: Tia Harrison
Email: tia.harrison@mail.utoronto.ca

---

# Overview

- Filtering (Wickham and Grolemund, 2017 Chapter 5, Timbers et al. 2021, Chapter 3.6) 
- Arranging (Wickham and Grolemund, 2017 Chapter 5)
- Selecting (Wickham and Grolemund, 2017 Chapter 5, Timbers et al. 2021, Chapter 3.5)
- The pipe (Wickham and Grolemund, 2017 Chapter 5 & 18; Timbers et al. 2021, Chapter 3.8)
- Mutating (Wickham and Grolemund, 2017 Chapter 5, Timbers et al. 2021, Chapter 3.7, 3.10)
- Summarising (Wickham and Grolemund, 2017 Chapter 5, Timbers et al. 2021, Chapter 3.9)
- Grouping (Wickham and Grolemund, 2017 Chapter 5)
- Cleaning (Alexander, 2022, Chapter 11)


---
# Take a look
```{r, echo=F}
ads_data <- haven::read_sav("ads_raw.sav")
```

```{r}
glimpse(ads_data)
```

---

# Filtering

Filtering allows us to select rows based on specific traits

```{r}
filter(ads_data, Duration__in_seconds_ < 100)
```

---

# Arranging

Arranging allows us to sort the order of the table by a certain column

```{r}
arrange(ads_data, Duration__in_seconds_)
```

---

# Selecting

Selecting allows us to pick certain columns

```{r}
select(ads_data, RecordedDate)
```

---
# Selecting

We can also remove columns

```{r}
select(ads_data, -Consent, -DistributionChannel)
```


---

# The pipe

So far, we have written our code like this:

```{r}
filter(ads_data, Duration__in_seconds_ < 100)
```

But what if we want to perform multiple operations in one go? 

---
# The pipe

We can use the pipe `%>%`, which passes what we wrote on the previous line into the next function as the first argument:

```{r}
ads_data %>%
  filter(Duration__in_seconds_ < 100) %>%
  arrange(Duration__in_seconds_) %>%
  select(RecordedDate, Duration__in_seconds_)
```

---
# The pipe

```{r, eval=FALSE}
ads_data %>%
  filter(Duration__in_seconds_ < 100) %>%
  arrange(Duration__in_seconds_) %>%
  select(RecordedDate, Duration__in_seconds_)
```

You can think of this like:

- Take the ADS data
- Filter so we only have the rows where the survey duration is less than 100 seconds
- Arrange so we go from lowest duration to highest
- Select only the date recorded and the duration

---

# Mutating

Mutating can be used to create new columns or change existing columns.

```{r}
ads_data <- ads_data %>%
  mutate(Birthyear_add_day = str_c(Birthyear, "07-01")) %>%
  mutate(Birthyear_add_day = as_datetime(Birthyear_add_day))
```

```{r echo=FALSE}
ads_data %>%
  select(EndDate, Birthyear, Birthyear_add_day)
```

---
# Mutating

```{r, eval=FALSE}
ads_data %>%
  mutate(age = EndDate - Birthyear_add_day) 
```

```{r, echo=FALSE}
ads_data %>%
  mutate(age = EndDate - Birthyear_add_day) %>%
  select(EndDate, Birthyear, Birthyear_add_day, age) 
```

---

# Summary

```{r}
summary(ads_data)
```

---

# Pulling a variable for calculations

```{r}
ads_data %>%
  pull(Duration__in_seconds_)
```

---

# Using the pulled variable for descriptive statistics

Median

```{r}
ads_data %>%
  pull(Duration__in_seconds_) %>%
  median(na.rm = TRUE)
```

We have to tell the mean() function to disregard NAs by writing `na.rm = TRUE`

---
# Using the pulled variable for descriptive statistics

Mean

```{r}
ads_data %>%
  pull(Duration__in_seconds_) %>%
  mean(na.rm = TRUE)
```

---
# Using the pulled variable for descriptive statistics


Range can be calculated using the `range()` function.

```{r}
ads_data %>%
  pull(Duration__in_seconds_) %>%
  range(na.rm = TRUE)
```

Variance can be calculated using the `var()` function.

```{r}
ads_data %>%
  pull(Duration__in_seconds_) %>%
  var(na.rm = TRUE)
```

---
# Using the pulled variable for descriptive statistics

Standard Deviation can be calculated using the `sd()` function.

```{r}
ads_data %>%
  pull(Duration__in_seconds_) %>%
  sd(na.rm = TRUE)
```

---

# Summarise

```{r}
ads_data %>%
  summarise(mean_time = mean(Duration__in_seconds_, na.rm = TRUE),
            sd_time = sd(Duration__in_seconds_, na.rm = TRUE))
```

---

# Grouping

Before summarising, we can group by a categorical variable

```{r}
ads_data %>%
  group_by(Gender) %>%
  summarise(count = n(),
            mean_time = mean(Duration__in_seconds_, na.rm = TRUE),
            sd_time = sd(Duration__in_seconds_, na.rm = TRUE))
```

---

class: inverse, center, middle

# Manipulation application: data cleaning

---
# Data cleaning

```{r, include=F, echo=F, message=F, warning=F}
ces_2019_raw <- read_csv("ces_2019_raw.csv")

CES_data <- ces_2019_raw %>%
  filter(cps19_Q_TotalDuration < 60 * 60) %>%
  mutate(cps19_province = fct_recode(factor(cps19_province),
                                       "Alberta" = "14",
                                       "British Columbia" = "15", 
                                       "Manitoba" = "16",
                                       "New Brunswick" = "17",
                                       "Newfoundland and Labrador" = "18",
                                       "Nova Scotia" = "20",
                                       "Nunavut" = "21",
                                       "Ontario" = "22",
                                     "Northwest Territories" = "19",
                                     "Prince Edward Island" = "23",
                                     "Quebec" = "24",
                                     "Saskatchewan" = "25",
                                     "Yukon" = "26"),
         cps19_votechoice = fct_recode(factor(cps19_votechoice),
                                       "Liberal Party" = "1",
                                       "Conservative Party" = "2", 
                                       "NDP" = "3",
                                       "Bloc Québécois" = "4",
                                       "Green Party" = "5",
                                       "People's Party" = "6",
                                       "Another party" = "7",
                                       "Don't know/ Prefer not to answer" = "9"),
         cps19_fed_donate = fct_recode(factor(cps19_fed_donate),
                                       "Yes" = "1",
                                       "No" = "2", 
                                       "Don't know/ Prefer not to answer" = "3"),
         cps19_vote_2015 = fct_recode(factor(cps19_vote_2015),
                                       "Liberal Party" = "1",
                                       "Conservative Party" = "2", 
                                       "NDP" = "3",
                                       "Bloc Québécois" = "4",
                                       "Green Party" = "5",
                                       "Another party" = "6",
                                       "Don't know/ Prefer not to answer" = "7"),
         cps19_news_cons = fct_recode(factor(cps19_news_cons),
                                       "0 minutes" = "1",
                                       "1-10 minutes" = "2", 
                                       "11-30 minutes" = "3",
                                       "31-60 minutes" = "4",
                                       "Between 1 and 2 hours" = "5",
                                       "More than 2 hours" = "6",
                                       "Don't know/ Prefer not to answer" = "7"),
         cps19_gender = fct_recode(factor(cps19_gender),
                                       "M" = "1",
                                       "F" = "2", 
                                       "NB" = "3"),
         cps19_spend_educ = fct_recode(factor(cps19_spend_educ),
                                       "Spend less" = "1",
                                       "Spend about the same as now" = "2", 
                                       "Spend more" = "3",
                                   "Don't know/ Prefer not to answer" = "4"),
         cps19_prov_gov_sat = fct_recode(factor(cps19_prov_gov_sat),
                                     "Very satisfied" = "1",
                                       "Fairly satisfied" = "2",
                                       "Not very satisfied" = "3",
                                       "Not at all satisfied" = "4",
                                       "Don't know/prefer not to answer" = "5"),
         cps19_prov_id = fct_recode(factor(cps19_prov_id),
                                "Liberal" = "281",
                                  "NDP" = "282",
                                  "Green" = "283",
                                  "Progressive Conservative" = "292",
                                  "Another party" = "295",
                                  "None" = "296",
                                  "Don't know/prefer not to answer" = "297"),
         cps19_union = fct_recode(factor(cps19_union),
                                       "Yes" = "1",
                                       "No" = "2", 
                                       "Don't know/ Prefer not to answer" = "3"),
         cps19_bornin_canada = fct_recode(factor(cps19_bornin_canada),
                                       "Yes" = "1",
                                       "No" = "2", 
                                       "Don't know/ Prefer not to answer" = "3"),
         cps19_income_number = ifelse(cps19_income_number > 1000000,
                                      NA,
                                      cps19_income_number),
         cps19_turnout_2015 = fct_recode(factor(cps19_turnout_2015),
                                     "Yes" = "1",
                                       "No" = "2",
                                       "Not eligible to vote in last election" = "3",
                                        "Don't know/ Prefer not to answer" = "4"),
         cps19_household = ifelse(cps19_household < 100, cps19_household, NA)
         )
```

Graphing year of birth shows that it goes from 1 to about 80.

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
ces_2019_raw %>%
  ggplot(aes(x = cps19_yob)) +
  geom_histogram()
```

---
# Data cleaning

The codebook says that a value of 1 corresponds to a birth year of 1920, value of 2 to a birth year of 1921, and so on. We can create a new variable that reads more intuitively.

```{r}
CES_data <- ces_2019_raw %>%
  mutate(cps19_yob_fix = cps19_yob + 1919)
```

---
# Data cleaning

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot(aes(x = cps19_yob_fix)) +
  geom_histogram()
```

Better!

---

# Add a variable for age

Now that we have an accurate birth year, maybe we would like to have the age of the individual as well.

```{r}
CES_data <- CES_data %>%
  mutate(age = 2019 - cps19_yob_fix)
```

---

# Add a variable for age

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot(aes(x = age)) +
  geom_histogram()
```

---

# Recoding the gender variable

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot(aes(x = cps19_gender)) + 
  geom_bar()
```

---

# Recoding the gender variable

```{r}
CES_data <- CES_data %>%
  mutate(cps19_gender_fix = factor(cps19_gender)) %>%
  mutate(cps19_gender_fix = fct_recode(cps19_gender_fix,
                                       "M" = "1",
                                       "F" = "2", 
                                       "NB" = "3"))
```

---
# Recoding the gender variable

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot(aes(x = cps19_gender_fix)) + 
  geom_bar()
```

---

# Fixing household counts

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  filter(cps19_household > 10) %>%
  arrange(-cps19_household) %>%
  pull(cps19_household)
```

---
# Fixing household counts

```{r}
CES_data <- CES_data %>%
  mutate(cps19_household = ifelse(cps19_household > 15, 
                                  NA, 
                                  cps19_household))

CES_data %>%
  filter(cps19_household > 10) %>%
  pull(cps19_household)
```

---

# Fixing income

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  filter(cps19_income_number > 1000000) %>%
  arrange(-cps19_income_number) %>%
  pull(cps19_income_number)
```

---
# Fixing income

```{r}
CES_data <- CES_data %>%
  mutate(cps19_income_number = ifelse(cps19_income_number >= 1000000000, 
                                  NA, 
                                  cps19_income_number))

CES_data %>%
  filter(cps19_income_number > 1000000) %>%
  pull(cps19_income_number)
```

---

class: inverse, center, middle

# Manipulation application: Summarising data

---
# Summarising data

```{r, include=F, echo=F, message=F, warning=F}
ces_2019_raw <- read_csv("ces_2019_raw.csv")

CES_data <- ces_2019_raw %>%
  filter(cps19_Q_TotalDuration < 60 * 60) %>%
  mutate(cps19_province = fct_recode(factor(cps19_province),
                                       "Alberta" = "14",
                                       "British Columbia" = "15", 
                                       "Manitoba" = "16",
                                       "New Brunswick" = "17",
                                       "Newfoundland and Labrador" = "18",
                                       "Nova Scotia" = "20",
                                       "Nunavut" = "21",
                                       "Ontario" = "22",
                                     "Northwest Territories" = "19",
                                     "Prince Edward Island" = "23",
                                     "Quebec" = "24",
                                     "Saskatchewan" = "25",
                                     "Yukon" = "26"),
         cps19_votechoice = fct_recode(factor(cps19_votechoice),
                                       "Liberal Party" = "1",
                                       "Conservative Party" = "2", 
                                       "NDP" = "3",
                                       "Bloc Québécois" = "4",
                                       "Green Party" = "5",
                                       "People's Party" = "6",
                                       "Another party" = "7",
                                       "Don't know/ Prefer not to answer" = "9"),
         cps19_fed_donate = fct_recode(factor(cps19_fed_donate),
                                       "Yes" = "1",
                                       "No" = "2", 
                                       "Don't know/ Prefer not to answer" = "3"),
         cps19_vote_2015 = fct_recode(factor(cps19_vote_2015),
                                       "Liberal Party" = "1",
                                       "Conservative Party" = "2", 
                                       "NDP" = "3",
                                       "Bloc Québécois" = "4",
                                       "Green Party" = "5",
                                       "Another party" = "6",
                                       "Don't know/ Prefer not to answer" = "7"),
         cps19_news_cons = fct_recode(factor(cps19_news_cons),
                                       "0 minutes" = "1",
                                       "1-10 minutes" = "2", 
                                       "11-30 minutes" = "3",
                                       "31-60 minutes" = "4",
                                       "Between 1 and 2 hours" = "5",
                                       "More than 2 hours" = "6",
                                       "Don't know/ Prefer not to answer" = "7"),
         cps19_gender = fct_recode(factor(cps19_gender),
                                       "M" = "1",
                                       "F" = "2", 
                                       "NB" = "3"),
         cps19_spend_educ = fct_recode(factor(cps19_spend_educ),
                                       "Spend less" = "1",
                                       "Spend about the same as now" = "2", 
                                       "Spend more" = "3",
                                   "Don't know/ Prefer not to answer" = "4"),
         cps19_prov_gov_sat = fct_recode(factor(cps19_prov_gov_sat),
                                     "Very satisfied" = "1",
                                       "Fairly satisfied" = "2",
                                       "Not very satisfied" = "3",
                                       "Not at all satisfied" = "4",
                                       "Don't know/prefer not to answer" = "5"),
         cps19_prov_id = fct_recode(factor(cps19_prov_id),
                                "Liberal" = "281",
                                  "NDP" = "282",
                                  "Green" = "283",
                                  "Progressive Conservative" = "292",
                                  "Another party" = "295",
                                  "None" = "296",
                                  "Don't know/prefer not to answer" = "297"),
         cps19_union = fct_recode(factor(cps19_union),
                                       "Yes" = "1",
                                       "No" = "2", 
                                       "Don't know/ Prefer not to answer" = "3"),
         cps19_bornin_canada = fct_recode(factor(cps19_bornin_canada),
                                       "Yes" = "1",
                                       "No" = "2", 
                                       "Don't know/ Prefer not to answer" = "3"),
         cps19_income_number = ifelse(cps19_income_number > 1000000,
                                      NA,
                                      cps19_income_number),
         cps19_turnout_2015 = fct_recode(factor(cps19_turnout_2015),
                                     "Yes" = "1",
                                       "No" = "2",
                                       "Not eligible to vote in last election" = "3",
                                        "Don't know/ Prefer not to answer" = "4"),
         cps19_household = ifelse(cps19_household < 100, cps19_household, NA)
         )
```

First we can select only data for Ontario using `filter()`:

```{r}
CES_data %>%
  filter(cps19_province == "Ontario")
```

---
# Summarising data

We don't need to be dealing with all the columns. We can specifically select the ones we want using `select()`:

"How satisfied are you with the performance of your provincial government under ${e://Field/premier}?", "In provincial politics, do you usually think of yourself as a:", and income.

```{r}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number)
```

---
# Summarising data
Now that our data looks like what we would like it to, we can start creating a summary table. Since we have the income for each participant, we can look at median incomes. We also want to know how many participants are in each category.

First, we can group the data by provincial political self-ID. To do this, we use `group_by()` to group the data and `summarise()` to produce values for each group we have created. We will start with calculating the `median()` for the incomes. We can add multiple arguments to the `summarise()` argument. `n()` adds a count for each group.

---
# Summarising data
```{r}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  group_by(cps19_prov_gov_sat) %>%
  summarise(median_income = median(cps19_income_number,
                                   na.rm = TRUE),
            count = n())
```

---
# Grouping
In our table, the satisfaction ratings are ordered alphabetically. We would like them to be ordered logically. We can do this by ordering the factor variable.

```{r, eval=F}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  mutate(cps19_prov_id = factor(cps19_prov_id,
                                     levels = c("Liberal",
                                                "Progressive Conservative",
                                                "NDP",
                                                "Green",
                                                "Another party",
                                                "None",
                                                "Don't know/prefer not to answer")))
```

---
# Grouping

```{r, echo=F}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  mutate(cps19_prov_id = factor(cps19_prov_id,
                                     levels = c("Liberal",
                                                "Progressive Conservative",
                                                "NDP",
                                                "Green",
                                                "Another party",
                                                "None",
                                                "Don't know/prefer not to answer")))
```

---
# Grouping
And combine this with our table from before:

```{r}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  mutate(cps19_prov_gov_sat = factor(cps19_prov_gov_sat,
                                     levels = c("Not at all satisfied",
                                                "Not very satisfied",
                                                "Fairly satisfied",
                                                "Very satisfied",
                                                "Don't know/prefer not to answer"))) %>%
  group_by(cps19_prov_gov_sat) %>%
  summarise(median_income = median(cps19_income_number,
                                   na.rm = TRUE),
            count = n())
```

---
# Grouping
What happens if we group by political identification instead?

```{r}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  group_by(cps19_prov_id) %>%
  summarise(median_income = median(cps19_income_number,
                                   na.rm = TRUE),
            count = n())
```

---
# Grouping
We could order the parties in a way that makes more sense:

```{r, eval=F}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  mutate(cps19_prov_id = factor(cps19_prov_id,
                                     levels = c("Liberal",
                                                "Progressive Conservative",
                                                "NDP",
                                                "Green",
                                                "Another party",
                                                "None",
                                                "Don't know/prefer not to answer"))) %>%
  group_by(cps19_prov_id) %>%
  summarise(median_income = median(cps19_income_number,
                                   na.rm = TRUE),
            count = n())
```

---
# Grouping

```{r, echo=F}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  mutate(cps19_prov_id = factor(cps19_prov_id,
                                     levels = c("Liberal",
                                                "Progressive Conservative",
                                                "NDP",
                                                "Green",
                                                "Another party",
                                                "None",
                                                "Don't know/prefer not to answer"))) %>%
  group_by(cps19_prov_id) %>%
  summarise(median_income = median(cps19_income_number,
                                   na.rm = TRUE),
            count = n())
```

---
# Grouping
Or we could sort by median income. We can do that using `arrange()`:

```{r}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  group_by(cps19_prov_id) %>%
  summarise(median_income = median(cps19_income_number,
                                   na.rm = TRUE),
            count = n()) %>%
  arrange(-median_income)
```

---
# Grouping
`group_by()` can also have multiple arguments, so we can group by `cps19_prov_gov_sat` and `cps19_prov_id` at the same time:

```{r}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  mutate(cps19_prov_id = factor(cps19_prov_id,
                                     levels = c("Liberal",
                                                "Progressive Conservative",
                                                "NDP",
                                                "Green",
                                                "Another party",
                                                "None",
                                                "Don't know/prefer not to answer"))) %>%
  mutate(cps19_prov_gov_sat = factor(cps19_prov_gov_sat,
                                     levels = c("Not at all satisfied",
                                                "Not very satisfied",
                                                "Fairly satisfied",
                                                "Very satisfied",
                                                "Don't know/prefer not to answer"))) %>%
  group_by(cps19_prov_gov_sat, cps19_prov_id) %>%
  summarise(median_income = median(cps19_income_number,
                                   na.rm = TRUE))
```

---
# Grouping
This table is less easy to read, though. `spread()` can make a table that is wide rather than long. We specify the `key`, the variable that will become our column names, and the `value`, which will become the values in those columns:

```{r, eval=F}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  mutate(cps19_prov_id = factor(cps19_prov_id,
                                     levels = c("Liberal",
                                                "Progressive Conservative",
                                                "NDP",
                                                "Green",
                                                "Another party",
                                                "None",
                                                "Don't know/prefer not to answer"))) %>%
  mutate(cps19_prov_gov_sat = factor(cps19_prov_gov_sat,
                                     levels = c("Not at all satisfied",
                                                "Not very satisfied",
                                                "Fairly satisfied",
                                                "Very satisfied",
                                                "Don't know/prefer not to answer"))) %>%
  group_by(cps19_prov_gov_sat, cps19_prov_id) %>%
  summarise(median_income = median(cps19_income_number,
                                   na.rm = TRUE)) %>%
  spread(key = cps19_prov_gov_sat,
         value = median_income)
```


---
# Grouping
```{r, echo=F}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  mutate(cps19_prov_id = factor(cps19_prov_id,
                                     levels = c("Liberal",
                                                "Progressive Conservative",
                                                "NDP",
                                                "Green",
                                                "Another party",
                                                "None",
                                                "Don't know/prefer not to answer"))) %>%
  mutate(cps19_prov_gov_sat = factor(cps19_prov_gov_sat,
                                     levels = c("Not at all satisfied",
                                                "Not very satisfied",
                                                "Fairly satisfied",
                                                "Very satisfied",
                                                "Don't know/prefer not to answer"))) %>%
  group_by(cps19_prov_gov_sat, cps19_prov_id) %>%
  summarise(median_income = median(cps19_income_number,
                                   na.rm = TRUE)) %>%
  spread(key = cps19_prov_gov_sat,
         value = median_income)
```


---

class: inverse, center, middle

# Exercises

---
# Exercises

1. Filter the rows in the CES_data dataset where the survey-taker is between 30 and 50 (cps19_age).
2. Filter the rows in the CES_data dataset where the survey-taker answered the cps19_votechoice question (i.e. the cps19_votechoice variable is not NA).
3. Select the variables cps19_age and cps19_province from the CES_data dataset.
4. Select all variables except cps19_province from the CES_data dataset.

---
# Exercises

1. Create a variable in the dataset CES_data that states if a person consumes news content or not (i.e. cps19_news_cons is equal to "0 minutes" or it is not).
2. Modify the variable cps19_income_number in the dataset CES_data so that it is measured in thousands (i.e. divide the income number by 1000).

---
# Exercises

1. Use the CES_data dataset. Group by cps19_votechoice. Find both the median and mean rating of Trudeau (cps19_lead_rating_23):
2. Use the CES_data dataset. Group by cps19_imm and cps19_spend_educ. Find the count for each group.

---
# Exercises

* 1 - Fix this error:

```{r, eval=F}
CES_data %>%
  summarise(mean = mean(cps19_age)) %>%
  group_by(cps19_gender)
```

* 2 - Fix this error:

```{r, eval=F}
CES_data %>%
  filter(cps19_vote_choice == "Green Party")
```
---
# Exercises

* 3 - Fix this error:

```{r, eval=F}
CES_data %>%
  mutate(cps19_fed_donate = factor(cps19_fed_donate,
                                     levels = c("Yes",
                                                "No",
                                                "Don't know/ Prefer not to answer"))
```

* 4 - Fix this error:

```{r, eval=F}
CES_data %>%
  select(cps19_province
         cps19_age
         cps19_gender)
```

---

class: inverse, center, middle

# Any questions?



