---
title: "Module 3: R"
subtitle: "Visualization"
author:
- "Instructor: Anjali Silva, PhD"
institute: "Data Sciences Institute, University of Toronto"
date: "2022"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  colors = c(
    red = "#f34213",
    purple = "#3e2f5b",
    orange = "#ff8811",
    green = "#136f63",
    white = "#FFFFFF"
  )
)

```

```{r, include=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)
library(kableExtra)
library(haven)
library(rgdal)
```

# Course Documents
* Visit: https://github.com/anjalisilva/IntroductionToR 

* All course material will be available via IntroductionToR GitHub repository (https://github.com/anjalisilva/IntroductionToR). Folder structure is as follows:
   * Lessons - All files: These folder contains all files.
   * **Lessons - Data only**: This folder contains data only.
   * **Lessons - Lesson Plans only**: This folder contains lesson plans only.
   * **Lessons - PDF only**: This folder contains slide PDFs only.
   * README - README file
   * .gitignore - Files to ignore specified by instructor

## Course Contacts
* Instructor: Anjali Silva
  Email: a.silva@utoronto.ca (Must use the subject line DSI-IntroR.   E.g., DSI-IntroR: Inquiry about Lecture I.)

* TA: Tia Harrison
Email: tia.harrison@mail.utoronto.ca

---

# Overview

- ggplot
- Initializing plots, specifying variables, and choosing chart types
- Customizing plots with labels, axes, color, size, multiple graph types, multiple plots, and overall look

(Wickham and Grolemund, 2017, Chapter 3; Healy, 2018, Chapter 3; Alexander, 2022, Chapter 6)

---

class: inverse, center, middle

# ggplot 

---
# ggplot
ggplot2 is a package that allows us to make graphics in R. It’s loaded with the tidyverse.

---
# Initializing a plot

the `ggplot()` function initializes the plot. In the arguments, you will identify the base of your plot. This includes:

- the data we want to graph from
- the aesthetics we will use

What doesn't go in the `ggplot()` function:

- the type of graph we want
- the way we want the axes to look
- the labels we want

```{r, eval=FALSE}
ggplot(data = my_data, mapping = aes())
```

---

# Aesthetic

In the `mapping` argument, we specific our aesthetic using `aes()`

This is where we indicate what variable we want for the x axis, the y axis, the color, or any other feature that the plot in question might have.

```{r, eval=FALSE}
aes(x = variable1,
    y = variable2,
    shape = variable3,
    color = variable4,
    size = variable5,
    fill = variable6)
```

---

# In combination

This takes `data` and passes it to ggplot, which initializes a plot using that data and specifies that variable1 will be represented on the x axis and variable2 on the y axis.

```{r, eval=FALSE}
data %>%
  ggplot(aes(x = variable1, y = variable2))
```

---

# Adding layers

After initializing, you still won't have a plot. You have to add layers -- which includes the type of plot you want, as well as tweaks and formatting specifics.

When we add layers to a ggplot object, we use the `+` between lines:

```{r, eval=FALSE}
data %>%
  ggplot(aes(x = variable1, y = variable2)) +
  # a layer +
  # another layer
```

---

# Geoms

Geom layers add types of plot. There are more than 40 geoms! Some common ones:

Bar plots

```{r eval=FALSE}
geom_bar()
```

Histograms

```{r eval=FALSE}
geom_histogram()
```

Scatterplots

```{r eval=FALSE}
geom_point()
```

---

# Examples

```{r, include=F, echo=F, message=F, warning=F}
ces_2019_raw <- read_csv("ces_2019_raw.csv")

CES_data <- ces_2019_raw %>%
  filter(cps19_Q_TotalDuration < 60 * 60) %>%
  mutate(cps19_province = fct_recode(factor(cps19_province),
                                       "Alberta" = "14",
                                       "British Columbia" = "15", 
                                       "Manitoba" = "16",
                                       "New Brunswick" = "17",
                                       "Newfoundland and Labrador" = "18",
                                       "Nova Scotia" = "20",
                                       "Nunavut" = "21",
                                       "Ontario" = "22",
                                     "Northwest Territories" = "19",
                                     "Prince Edward Island" = "23",
                                     "Quebec" = "24",
                                     "Saskatchewan" = "25",
                                     "Yukon" = "26"),
         cps19_votechoice = fct_recode(factor(cps19_votechoice),
                                       "Liberal Party" = "1",
                                       "Conservative Party" = "2", 
                                       "NDP" = "3",
                                       "Bloc Québécois" = "4",
                                       "Green Party" = "5",
                                       "People's Party" = "6",
                                       "Another party" = "7",
                                       "Don't know/ Prefer not to answer" = "9"),
         cps19_fed_donate = fct_recode(factor(cps19_fed_donate),
                                       "Yes" = "1",
                                       "No" = "2", 
                                       "Don't know/ Prefer not to answer" = "3"),
         cps19_vote_2015 = fct_recode(factor(cps19_vote_2015),
                                       "Liberal Party" = "1",
                                       "Conservative Party" = "2", 
                                       "NDP" = "3",
                                       "Bloc Québécois" = "4",
                                       "Green Party" = "5",
                                       "Another party" = "6",
                                       "Don't know/ Prefer not to answer" = "7"),
         cps19_news_cons = fct_recode(factor(cps19_news_cons),
                                       "0 minutes" = "1",
                                       "1-10 minutes" = "2", 
                                       "11-30 minutes" = "3",
                                       "31-60 minutes" = "4",
                                       "Between 1 and 2 hours" = "5",
                                       "More than 2 hours" = "6",
                                       "Don't know/ Prefer not to answer" = "7"),
         cps19_gender = fct_recode(factor(cps19_gender),
                                       "M" = "1",
                                       "F" = "2", 
                                       "NB" = "3"),
         cps19_spend_educ = fct_recode(factor(cps19_spend_educ),
                                       "Spend less" = "1",
                                       "Spend about the same as now" = "2", 
                                       "Spend more" = "3",
                                   "Don't know/ Prefer not to answer" = "4"),
         cps19_prov_gov_sat = fct_recode(factor(cps19_prov_gov_sat),
                                     "Very satisfied" = "1",
                                       "Fairly satisfied" = "2",
                                       "Not very satisfied" = "3",
                                       "Not at all satisfied" = "4",
                                       "Don't know/prefer not to answer" = "5"),
         cps19_prov_id = fct_recode(factor(cps19_prov_id),
                                "Liberal" = "281",
                                  "NDP" = "282",
                                  "Green" = "283",
                                  "Progressive Conservative" = "292",
                                  "Another party" = "295",
                                  "None" = "296",
                                  "Don't know/prefer not to answer" = "297"),
         cps19_union = fct_recode(factor(cps19_union),
                                       "Yes" = "1",
                                       "No" = "2", 
                                       "Don't know/ Prefer not to answer" = "3"),
         cps19_bornin_canada = fct_recode(factor(cps19_bornin_canada),
                                       "Yes" = "1",
                                       "No" = "2", 
                                       "Don't know/ Prefer not to answer" = "3"),
         cps19_income_number = ifelse(cps19_income_number > 1000000,
                                      NA,
                                      cps19_income_number),
         cps19_turnout_2015 = fct_recode(factor(cps19_turnout_2015),
                                     "Yes" = "1",
                                       "No" = "2",
                                       "Not eligible to vote in last election" = "3",
                                        "Don't know/ Prefer not to answer" = "4"),
         cps19_household = ifelse(cps19_household < 100, cps19_household, NA)
         )
```

---

# Bar plot

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot(aes(x = cps19_votechoice, )) +
  geom_bar()
```

---

# Options for barplots

These are the defaults in a geom_bar:

```{r, eval=FALSE}
geom_bar(
  stat = "count", # can change to "prop" for proportion
  position = "stack", # can change to "dodge" or "fill"
  width = NULL, # can put a bar width here
  na.rm = FALSE, # can remove NAs
  orientation = NA,  # can specify "x" or "y"
  show.legend = NA # can add or remove a legend
)
```


---

# Histogram

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  mutate(age = 90 - cps19_yob) %>%
  ggplot(aes(x = cps19_age)) +
  geom_histogram()
```

---

# Options for histograms

```{r, eval=FALSE}
geom_histogram(
  stat = "bin", # can change to "count"
  position = "stack", # can change to "identity", "dodge", or "fill"
  binwidth = NULL, # can specify the range of each bin
  bins = NULL, # can specify the number of bins
  na.rm = FALSE, # can tell it to ignore NAs
  orientation = NA, # can specify "x" or "y"
  show.legend = NA # can add or remove a legend
)
```

---

# Scatter plot

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  mutate(age = 90 - cps19_yob) %>%
  ggplot(aes(x = cps19_age, y = cps19_household)) +
  geom_point()
```

---

# Options for scatterplots

```{r, eval=FALSE}
geom_point(
  position = "identity", # can change to "jitter"
  na.rm = FALSE, # can ignore NAs
  show.legend = NA # can add or remove a legend
)
```

---

# Exercises

Using different variables:

1. Make a barplot
2. Make a histogram
3. Make a scatterplot 

---

class: inverse, center, middle

# Customizing plots

---

# Labels

We can change the way that labels appear to improve the look and readability.

```{r, eval=FALSE}
labs(x = "X name",
     y = NULL,
     title = "Title")
```

---

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot(aes(x = cps19_gender)) + 
  geom_bar() +
  labs(x = "Self-identified gender",
       y = "Count",
       title = "Gender frequencies in CES data")
```

---

# Axes

How we change the axes depends on what types of variables we have.

The layers take the form: `scale_<which axis>_<what type of axis>()`. There are VERY many `scale_` options.

Similar layers can be added for x and y axes, as well as other graph features like color and size.

---

# Manipulate a continuous x axis

```{r, eval=FALSE}
scale_x_continuous(breaks = , # use a vector to specify locations
                   minor_breaks = , # also can be a vector
                   n.breaks = , # can specify the number of breaks
                   labels = , # change the labels on the breaks
                   limits = , # set the upper and lower limits
                   position = # "left", "right", "top", "bottom")
```

---

# Manipulate a discrete x axis 

```{r, eval=FALSE}
scale_x_discrete(breaks = , # character vector of breaks
                  limits = , # set possible values for scale
                  drop = , # TRUE or FALSE to drop unused factor levels
                  labels = , # change the labels on the breaks
                  psiition = # "left", "right", "top", "bottom")
```

---

# Fill

We can change the fill to represent a variable:

```{r, eval=FALSE}
data %>%
  ggplot(aes(x = variable1,
             fill = variable2)) +
  geom_()
```

or to be a color of choice:

```{r, eval=FALSE}
data %>%
  ggplot(aes(x = variable1)) +
  geom_(fill = "my_color")
```

The difference is where the `fill =` is located. If it is in the aes(), then it will represent a variable. If it is not in the aes(), it will just change the look of the graph.

---

# Fill to represent a variable

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot(aes(x = cps19_age,
             fill = cps19_gender)) +
  geom_histogram()
```

---

# Changing fill to a specific color

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot(aes(x = cps19_age)) +
  geom_histogram(fill = "#6DAEDB")
```

---

# Color

For some geoms, you will need to change color rather than fill.

We can change the color to represent a variable:

```{r, eval=FALSE}
data %>%
  ggplot(aes(x = variable1,
             color = variable2)) +
  geom_()
```

or to be a color of choice:

```{r, eval=FALSE}
data %>%
  ggplot(aes(x = variable1)) +
  geom_(color = "my_color")
```

The difference is where the `color =` is located. If it is in the aes(), then it will represent a variable. If it is not in the aes(), it will just change the look of the graph.

---

# Color to represent a variable

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot(aes(x = cps19_age,
             y = cps19_household,
             color = cps19_gender)) +
  geom_smooth()
```

---

# Color for visual effect

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot(aes(x = cps19_age,
             y = cps19_household)) +
  geom_smooth(color = "#6DAEDB")
```

---

# Size

We can change the size to represent a variable:

```{r, eval=FALSE}
data %>%
  ggplot(aes(x = variable1,
             y = variable2,
             size = variable3)) +
  geom_()
```

or to be a color of choice:

```{r, eval=FALSE}
data %>%
  ggplot(aes(x = variable1,
             y = variable2)) +
  geom_(size = "my_size")
```

The difference is where the `fill =` is located. If it is in the aes(), then it will represent a variable. If it is not in the aes(), it will just change the look of the graph.

---

# Size to represent a variable

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot(aes(x = cps19_age,
             y = cps19_income_number,
             size = cps19_household)) +
  geom_point()
```

---

# Size for visual effect

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot(aes(x = cps19_age,
             y = cps19_income_number)) +
  geom_point(size = 15)
```

---

# Using multiple geoms

We can layer geoms on top of one another.

Geoms can either share their aes():

```{r, eval=FALSE}
data %>%
  ggplot(aes(x = variable1,
             y = variable2)) +
  geom_() +
  geom_()
```

or they can have their own:

```{r, eval=FALSE}
data %>%
  ggplot() +
  geom_(aes(x = variable1,
             y = variable2)) +
  geom_(aes(x = variable1,
             y = variable3))
```

---

# Geoms that share aesthetics

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot(aes(x = cps19_age,
             y = cps19_income_number)) +
  geom_point() +
  geom_smooth()
```

---

# Separate geom aesthetics

```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot() +
  geom_point(aes(x = cps19_age,
                 y = cps19_income_number)) +
  geom_line(aes(x = cps19_age,
                y = mean(cps19_income_number, na.rm = TRUE)))
```

---

# Facets

Facets give you side-by-side graphs for different categories.

```{r, eval=FALSE}
facet_wrap(facets = "variables you want to facet by") 
facet_grid(facets = "variables that you want to facet by")
```


---
# Facets
```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot(aes(x = cps19_age)) +
  geom_histogram() +
  facet_wrap(facets = "cps19_gender")
```

---

# Themes

Themes are added at the end. They control the overall look.

```{r, eval=FALSE}
theme_bw()
theme_classic()
theme_light()
theme_minimal()
```

---
# Themes
```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot(aes(x = cps19_age)) +
  geom_histogram() +
  facet_wrap(facets = "cps19_gender") +
  theme_minimal()
```

---
# Themes
```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=11}
CES_data %>%
  ggplot(aes(x = cps19_age)) +
  geom_histogram() +
  facet_wrap(facets = "cps19_gender") +
  theme_light()
```

---

# Example: Graphing CES data

We can make a bar graph representing the response to the question: "Which party do you think you will vote for?", named `cps19_votechoice`.

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_bar(aes(x = cps19_votechoice))
```

---
# Bar graphs
```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_bar(aes(x = cps19_votechoice))
```

---
# Bar graphs
You may want to use more short forms in the responses to make the graph more readable. To manipulate the axis labels, we use the `scale_x_discrete` function and specify what labels we want:

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_bar(aes(x = cps19_votechoice)) +
  scale_x_discrete(labels = c(
    "Liberal Party" = "Liberal",
    "Conservative Party" = "Conservative",
    "Bloc Québécois" = "Bloc",
    "Green Party" = "Green",
    "People's Party" = "PPC",
    "Another party" = "Other",
    "Don't know/ Prefer not to answer" = "DK/PNTA")
  )
```

---
# Bar graphs
```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_bar(aes(x = cps19_votechoice)) +
  scale_x_discrete(labels = c(
    "Liberal Party" = "Liberal",
    "Conservative Party" = "Conservative",
    "Bloc Québécois" = "Bloc",
    "Green Party" = "Green",
    "People's Party" = "PPC",
    "Another party" = "Other",
    "Don't know/ Prefer not to answer" = "DK/PNTA")
  )
```

---
# Bar graphs
We can also reorder bars from in a way that makes more sense. To do this, we take the variable and make it into a factor. Factors have a specific order, given in the `levels` argument:

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_bar(aes(x = factor(cps19_votechoice,
                          levels = c("Liberal Party",
                                     "Conservative Party",
                                     "Bloc Québécois",
                                     "Green Party",
                                     "People's Party",
                                     "Another party",
                                     "Don't know/ Prefer not to answer")))) +
  scale_x_discrete(labels = c(
    "Liberal Party" = "Liberal",
    "Conservative Party" = "Conservative",
    "Bloc Québécois" = "Bloc",
    "Green Party" = "Green",
    "People's Party" = "PPC",
    "Another party" = "Other",
    "Don't know/ Prefer not to answer" = "DK/PNTA")
  )
```

---
# Bar graphs
```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_bar(aes(x = factor(cps19_votechoice,
                          levels = c("Liberal Party",
                                     "Conservative Party",
                                     "Bloc Québécois",
                                     "Green Party",
                                     "People's Party",
                                     "Another party",
                                     "Don't know/ Prefer not to answer")))) +
  scale_x_discrete(labels = c(
    "Liberal Party" = "Liberal",
    "Conservative Party" = "Conservative",
    "Bloc Québécois" = "Bloc",
    "Green Party" = "Green",
    "People's Party" = "PPC",
    "Another party" = "Other",
    "Don't know/ Prefer not to answer" = "DK/PNTA")
  )
```

---
# Bar graphs
What if we want to compare the voting intentions between genders? We can use the `fill` argument in the `aes()` function to do that:

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_bar(aes(x = factor(cps19_votechoice,
                          levels = c("Liberal Party",
                                     "Conservative Party",
                                     "Bloc Québécois",
                                     "Green Party",
                                     "People's Party",
                                     "Another party",
                                     "Don't know/ Prefer not to answer")),
               fill = cps19_gender)) +
  scale_x_discrete(labels = c(
    "Liberal Party" = "Liberal",
    "Conservative Party" = "Conservative",
    "Bloc Québécois" = "Bloc",
    "Green Party" = "Green",
    "People's Party" = "PPC",
    "Another party" = "Other",
    "Don't know/ Prefer not to answer" = "DK/PNTA"
  ))
```

---
# Bar graphs
```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_bar(aes(x = factor(cps19_votechoice,
                          levels = c("Liberal Party",
                                     "Conservative Party",
                                     "Bloc Québécois",
                                     "Green Party",
                                     "People's Party",
                                     "Another party",
                                     "Don't know/ Prefer not to answer")),
               fill = cps19_gender)) +
  scale_x_discrete(labels = c(
    "Liberal Party" = "Liberal",
    "Conservative Party" = "Conservative",
    "Bloc Québécois" = "Bloc",
    "Green Party" = "Green",
    "People's Party" = "PPC",
    "Another party" = "Other",
    "Don't know/ Prefer not to answer" = "DK/PNTA"
  ))
```

---
# Bar graphs
If we don't want the bars to be stacked, we need to change the `position` argument in the `geom_bar()` function:

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_bar(aes(x = factor(cps19_votechoice,
                          levels = c("Liberal Party",
                                     "Conservative Party",
                                     "Bloc Québécois",
                                     "Green Party",
                                     "People's Party",
                                     "Another party",
                                     "Don't know/ Prefer not to answer")),
               fill = cps19_gender),
           position = "dodge") +
  scale_x_discrete(labels = c(
    "Liberal Party" = "Liberal",
    "Conservative Party" = "Conservative",
    "Bloc Québécois" = "Bloc",
    "Green Party" = "Green",
    "People's Party" = "PPC",
    "Another party" = "Other",
    "Don't know/ Prefer not to answer" = "DK/PNTA"
  ))
```

---
# Bar graphs
```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_bar(aes(x = factor(cps19_votechoice,
                          levels = c("Liberal Party",
                                     "Conservative Party",
                                     "Bloc Québécois",
                                     "Green Party",
                                     "People's Party",
                                     "Another party",
                                     "Don't know/ Prefer not to answer")),
               fill = cps19_gender),
           position = "dodge") +
  scale_x_discrete(labels = c(
    "Liberal Party" = "Liberal",
    "Conservative Party" = "Conservative",
    "Bloc Québécois" = "Bloc",
    "Green Party" = "Green",
    "People's Party" = "PPC",
    "Another party" = "Other",
    "Don't know/ Prefer not to answer" = "DK/PNTA"
  ))
```

---
# Bar graphs
We can change the widths of the bars as well:

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_bar(aes(x = factor(cps19_votechoice,
                          levels = c("Liberal Party",
                                     "Conservative Party",
                                     "Bloc Québécois",
                                     "Green Party",
                                     "People's Party",
                                     "Another party",
                                     "Don't know/ Prefer not to answer")),
               fill = cps19_gender),
           position = "dodge",
           width = 0.5) +
  scale_x_discrete(labels = c(
    "Liberal Party" = "Liberal",
    "Conservative Party" = "Conservative",
    "Bloc Québécois" = "Bloc",
    "Green Party" = "Green",
    "People's Party" = "PPC",
    "Another party" = "Other",
    "Don't know/ Prefer not to answer" = "DK/PNTA"
  ))
```

---
# Bar graphs
```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_bar(aes(x = factor(cps19_votechoice,
                          levels = c("Liberal Party",
                                     "Conservative Party",
                                     "Bloc Québécois",
                                     "Green Party",
                                     "People's Party",
                                     "Another party",
                                     "Don't know/ Prefer not to answer")),
               fill = cps19_gender),
           position = "dodge",
           width = 0.5) +
  scale_x_discrete(labels = c(
    "Liberal Party" = "Liberal",
    "Conservative Party" = "Conservative",
    "Bloc Québécois" = "Bloc",
    "Green Party" = "Green",
    "People's Party" = "PPC",
    "Another party" = "Other",
    "Don't know/ Prefer not to answer" = "DK/PNTA"
  ))
```

---
# Histograms
We can make a histogram representing the response to the question: "How do you feel about the federal political parties below? Set the slider to a number from 0 to 100, where 0 means you really dislike the party and 100 means you really like the party." and the Conservative Party, named `cps19_party_rating_24`.

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24))
```

---
# Histograms
```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24))
```

---
# Histograms
The histogram splits the range of values for Conservative Party rating into 30 bins automatically, but what if we want a different number of bins? We can change the `bins` argument in the `geom_histogram()` function:

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24),
                 bins = 10)
```

---
# Histograms
```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24),
                 bins = 10)
```

---
# Histograms
If we want to look at distributions of a variable in different groups, we can use something called faceting. To show what the Conservative Party ratings look like across different views on education spending, we can add the `facet_wrap` function:

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24),
                 bins = 10) +
  facet_wrap(facets = "cps19_spend_educ")
```

---
# Histograms
```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24),
                 bins = 10) +
  facet_wrap(facets = "cps19_spend_educ")
```

---
# Histograms
We can use the `nrow` arguments to say how many rows we want the facets to form:

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24),
                 bins = 10) +
  facet_wrap(facets = "cps19_spend_educ",
             nrow = 1)
```

---
# Histograms
```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24),
                 bins = 10) +
  facet_wrap(facets = "cps19_spend_educ",
             nrow = 1)
```

---
# Histograms
The order of the facets could be better. We use the same `factor` and `levels` method as before:

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24),
                 bins = 10) +
  facet_wrap(~factor(cps19_spend_educ,
                     levels = c("Spend less",
                                "Spend about the same as now",
                                "Spend more",
                                "Don't know/ Prefer not to answer")),
             nrow = 1)
```

---
# Histograms
```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24),
                 bins = 10) +
  facet_wrap(~factor(cps19_spend_educ,
                     levels = c("Spend less",
                                "Spend about the same as now",
                                "Spend more",
                                "Don't know/ Prefer not to answer")),
             nrow = 1)
```

---
# Histograms
For any graph, we probably want better labels than the variable names. We add the `labels()` function to specific labels for the x-axis, the y-axis, and the title:

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24),
                 bins = 10) +
  facet_wrap(~factor(cps19_spend_educ,
                     levels = c("Spend less",
                                "Spend about the same as now",
                                "Spend more",
                                "Don't know/ Prefer not to answer")),
             nrow = 1) +
  labs(x = "Conservative Party rating",
       y = "Count",
       title = "Views on education spending and the Conservative Party")
```

---
# Histograms
```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24),
                 bins = 10) +
  facet_wrap(~factor(cps19_spend_educ,
                     levels = c("Spend less",
                                "Spend about the same as now",
                                "Spend more",
                                "Don't know/ Prefer not to answer")),
             nrow = 1) +
  labs(x = "Conservative Party rating",
       y = "Count",
       title = "Views on education spending and the Conservative Party")
```

---
# Histograms
We can widen the graph by editing the code chunk, specifying `fig.height` and `fig.width`:

```{r, fig.height = 5, fig.width = 10, eval=F}
CES_data %>%
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24),
                 bins = 10) +
  facet_wrap(~factor(cps19_spend_educ,
                     levels = c("Spend less",
                                "Spend about the same as now",
                                "Spend more",
                                "Don't know/ Prefer not to answer")),
             nrow = 1) +
  labs(x = "Conservative Party rating",
       y = "Count",
       title = "Views on education spending and the Conservative Party")
```

---
# Histograms
```{r, fig.height = 5, fig.width = 10, echo=F}
CES_data %>%
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24),
                 bins = 10) +
  facet_wrap(~factor(cps19_spend_educ,
                     levels = c("Spend less",
                                "Spend about the same as now",
                                "Spend more",
                                "Don't know/ Prefer not to answer")),
             nrow = 1) +
  labs(x = "Conservative Party rating",
       y = "Count",
       title = "Views on education spending and the Conservative Party")
```

---
# Scatterplots
We can make a scatterplot representing the relationship between the ages of the survey-takers and the time they spent on the survey, named `cps19_age` and `cps19_Q_TotalDuration`, using `geom_point()`:

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_point(aes(x = cps19_age,
             y = cps19_Q_TotalDuration))
```

---
# Scatterplots
```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_point(aes(x = cps19_age,
             y = cps19_Q_TotalDuration))
```

---
# Scatterplots
There are some very large values for time spent on survey that makes it hard to see the rest. We can look at only the values in between 0 seconds and 3600 seconds, or 1 hour. Those that took longer are considered to be 'inattentive'.

One way to do this is to set limits on the axis using `scale_y_continuous()`:

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_point(aes(x = cps19_age,
             y = cps19_Q_TotalDuration)) +
  scale_y_continuous(limits = c(0,3600))
```

---
# Scatterplots
```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_point(aes(x = cps19_age,
             y = cps19_Q_TotalDuration)) +
  scale_y_continuous(limits = c(0,3600))
```

---
# Scatterplots
Many points sitting exactly on top of each other, like what's happening with age, makes it hard to read the graph. If we add the arguement `position = "jitter"` to the `geom_point()` function, ggplot will slightly separate points that are in exactly the same spot:

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_point(aes(x = cps19_age,
             y = cps19_Q_TotalDuration),
         position = "jitter") +
  scale_y_continuous(limits = c(0,3600))
```

---
# Scatterplots
Many points sitting exactly on top of each other, like what's happening with age, makes it hard to read the graph. If we add the arguement `position = "jitter"` to the `geom_point()` function, ggplot will slightly separate points that are in exactly the same spot:

```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_point(aes(x = cps19_age,
             y = cps19_Q_TotalDuration),
         position = "jitter") +
  scale_y_continuous(limits = c(0,3600))
```

---
# Scatterplots
Now it's just formed a cloud. To get separation, we can decrease the size of the individual points with the `size` argument in `geom_point()`:

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_point(aes(x = cps19_age,
             y = cps19_Q_TotalDuration),
         position = "jitter",
         size = 0.1) +
  scale_y_continuous(limits = c(0,3600))
```

---
# Scatterplots
```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_point(aes(x = cps19_age,
             y = cps19_Q_TotalDuration),
         position = "jitter",
         size = 0.1) +
  scale_y_continuous(limits = c(0,3600))
```

---
# Scatterplots
Again, we can add labels:

```{r, eval=F}
CES_data %>%
  ggplot() +
  geom_point(aes(x = cps19_age,
             y = cps19_Q_TotalDuration),
         position = "jitter",
         size = 0.1) +
  scale_y_continuous(limits = c(0,3600)) +
  labs(x = "Age",
       y = "Seconds spent on survey",
       title = "Age and time to complete survey amoung attentive survey-takers")
```

---
# Scatterplots
```{r, echo=F}
CES_data %>%
  ggplot() +
  geom_point(aes(x = cps19_age,
             y = cps19_Q_TotalDuration),
         position = "jitter",
         size = 0.1) +
  scale_y_continuous(limits = c(0,3600)) +
  labs(x = "Age",
       y = "Seconds spent on survey",
       title = "Age and time to complete survey amoung attentive survey-takers")
```

---

class: inverse, center, middle

# Exercises

---
# Exercises
Take your:

1. barplots,
2. histograms, and
3. scatterplot from before 

and customize them. Try to add each different customization to at least one plot.

---

class: inverse, center, middle

# Any questions?

